Type theories as quotient inductive-recursive types

### 2025-01-24 tex/main.tex
Fix the argument for the coherence: telescope substitution, multiple lifting, and substitution with lifted substitution need to be proved altogether

\title{Type theories as quotient inductive-recursive types}
\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC [funding].}
\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
\authorrunning{L.-T.~Chen and T.-C.~Tsai}
\Copyright{Liang-Ting Chen and Tzu-Chun Tsai}
\ccsdesc[500]{Theory of computation~Type theory}
\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families}
\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

@@

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}

@@

\section{Metatheory and formalisation}

@@

\section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
\LT{
\begin{enumerate}
  \item Goal: no transports in the definition.
  \item Motivation: make type substitution definitional so that we do not have to apply transport along structural rules for types explicitly.
  \item Develop the quotient inductive-inductive-recursive definition of type theory step by step (which needs to be locally confluent and terminating).
    \begin{enumerate}
      \item parallel substitution: type substitution
      \item the type of elements of $A$: term substitution needs to be split into two definitional substitution and explicit term substitution to maintain local confluence.
      \item $\Pi$-type: because of the substitution lifting $\sigma^+ \defeq (\pi_1\idS, \pi_2\idS)$ is used we need to introduce a definitional lifting to maintain local confluence.
        We prefer categorical combinator and avoid using $\left< t \right>$ in the definition, which would introduce transports in the definition.
      \item Other type formers can be introduced as usual (as long as we use categorical combinators).
    \end{enumerate}
\end{enumerate}
}

We begin with the definition of (parallel) substitution calculus~\cite{Martin-Lof1992} as a quotient inductive-inductive type, emphasising how the use of transports in its very definition complicates formal reasoning.
Then we introduce its quotient inductive-inductive-recursive counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning and we develop a type theory with a universe (\cref{subsec:SC+U}), $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).

@@

Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types under some context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from the domain $\Gamma$ and the codomain $\Delta$, and a type $\Tm\;\Gamma\;A$ of terms under a context $\Gamma$ and its type $A$. 
These types amount to the following types indexed by types being defined (hence inductive-inductive):
\begin{alignat*}{3}
  \Ctx   & : \Set                   \\
  \Ty    & : \Ctx \to \Set          \\
  \Sub   & : \Ctx \to \Ctx \to \Set \\
  \Tm    & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{alignat*}
The type $\Ctx$ has two constructors $\emptyctx$ for the empty context and $\blank,\blank$ for context extension:
\begin{alignat*}{3}
  \emptyctx & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
where the context extension $\blank,\blank$ requires that the type $\Ty$ indexed by a context $\Gamma$, underscoring the nature of inductive-inductive definition.

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma.
\end{alignat*}
where $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.

@@

Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitution extensions $\sigma, t$ by some term~$t$:
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
\end{alignat*}
Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
As substitution calculus is the initial cwf with a constant type $\UU$, substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$
\begin{alignat*}{3}
  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
\end{alignat*}
satisfying certain laws (introduced later).
Context comprehension is given by projections from \emph{non-empty} substitutions $\Sub\;\Gamma\;(\Delta, A)$:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
\end{alignat*}
where $\pi_1\,\sigma$ and $\pi_2\,\sigma$ can be intuitively understood as the \emph{tail} and the \emph{head} of a non-empty substitution $\sigma : \Sub\;\Gamma\;(\Delta, A)$ respectively.
Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
Finally, we also have term substitution $[\sigma]\,t$ whose type is $[\sigma]\,A$ for a term $t : \Tm\;\Delta\;A$
\begin{alignat*}{3}
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A)
\end{alignat*}
where the symbol $[\blank]\blank$ is overloaded.

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for a simple fact below.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following equality apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to eliminate that $\transp$ to derive the right hand side.

@@

To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor cannot compute so that type mismatches occur after substitution.

One possibility is to define substitution calculus as a QII\emph{R} type where type substitution, as a simultaneously defined recursion, does compute.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
Note that this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.}
Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
This definition fails to eliminate any transport in the definition of substitution calculus.

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
By design, definitions by pattern matching will be translated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching, so only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
\LT[noinline]{How about \Coq and \Lean?}
Using the first-match semantics, the above definition amounts to the following definition:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
\end{alignat*}
where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\begin{alignat*}{3}
[ \emptysub ]_{\Ty}\;\UU         & \reduce \UU \\
[ \sigma, t ]_{\Ty}\;\UU         & \reduce \UU \\
[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \reduce \UU,
\end{alignat*}
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional on \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.

@@

Yet, as we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent.
This boils down to that $[\blank]_{\Ty}\blank$ preserves all equality constructors.
\LT{Should we mention general condition for coherence?} 
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold (propositionally).
  The first two identities of the first row hold definitionally by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
For a QII definition \cite{Altenkirch2016a}, this extension amounts to adding the following constructors 
\begin{alignat*}{3}
  \UU   & : \implicit{\Gamma} && \Ty\, \Gamma \\
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}})
\end{alignat*}
where the type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.

>>>>

\title{Type theories as quotient inductive-recursive types}
\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.}
\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
\authorrunning{L.-T.~Chen and T.-C.~Tsai}
\Copyright{Liang-Ting Chen and Tzu-Chun Tsai}
\ccsdesc[500]{Theory of computation~Type theory}
\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families}
\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

@@

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}

@@

\section{Metatheory and formalisation} \label{sec:meta}

@@

\section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
\LT{
\begin{enumerate}
  \item Goal: remove transports in the definition so that less transports during reasoning.
  \item Motivation: make type substitution definitional so that we do not have to apply transport along structural rules for types explicitly.
  \item Develop the quotient inductive-inductive-recursive definition of type theory step by step (which needs to be locally confluent and terminating).
    \begin{enumerate}
      \item parallel substitution: type substitution
      \item the type of elements of $A$: term substitution needs to be split into two definitional substitution and explicit term substitution to maintain local confluence.
      \item $\Pi$-type: because of the substitution lifting $\sigma^+ \defeq (\pi_1\idS, \pi_2\idS)$ is used we need to introduce a definitional lifting to maintain local confluence.
        We prefer categorical combinator and avoid using $\left< t \right>$ in the definition, which would introduce transports in the definition.
      \item Other type formers can be introduced as usual (as long as we use categorical combinators).
    \end{enumerate}
\end{enumerate}
}

We begin with the definition of (parallel) substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
Then we introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning.
Then, we develop a type theory with a universe (\cref{subsec:SC+U}), $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).

@@

Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types under some context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from the domain $\Gamma$ and the codomain $\Delta$, and a type $\Tm\;\Gamma\;A$ of terms under a context $\Gamma$ and its type $A$. 
These types amount to the following types indexed by types being defined (hence inductive-inductive):
\begin{alignat*}{3}
  \Ctx   & : \Set                   \\
  \Ty    & : \Ctx \to \Set          \\
  \Sub   & : \Ctx \to \Ctx \to \Set \\
  \Tm    & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{alignat*}
The type $\Ctx$ has two constructors $\emptyctx$ for the empty context and $\blank,\blank$ for context extension:
\begin{alignat*}{3}
  \emptyctx & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
where the context extension $\blank,\blank$ requires that $\Ty$ is indexed by a context $\Gamma$, underscoring the nature of inductive-inductive definition.

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
\end{alignat*}
where $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.

@@

Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitution extensions $\sigma, t$ by some term~$t$:
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
\end{alignat*}
Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
As substitution calculus is the initial cwf with a constant type $\UU$, substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$
\begin{alignat*}{3}
  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
\end{alignat*}
satisfying certain laws (introduced later).
Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$
for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
\end{alignat*}
respectively.
Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution:
\begin{alignat*}{3}
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A).
\end{alignat*}
such that $[\sigma]\,t$ is of $\Tm\;\Gamma\;([\sigma]\,A)$.
Note that $[\blank]\blank$ is overloaded for type and term substitution.

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for a simple fact below.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following equality apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.

@@

To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor does not compute so that type mismatches occur after substitution.

One possibility is to define substitution calculus as a QII`R' type where type substitution, as a simultaneously defined recursion, does compute.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
Note that this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.}
Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional.
This definition fails to remove any transport in the definition of substitution calculus.

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} or an explicit construct for pattern matching.
In this case, only the eliminator for $\Ty\;\Delta$ will be considered in the above definition.
\LT[noinline]{How about \Coq and \Lean?}
Using the first-match semantics, the above definition amounts to the following definition:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
\end{alignat*}
where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\begin{alignat*}{3}
[ \emptysub ]_{\Ty}\;\UU         & \reduce \UU \\
[ \sigma, t ]_{\Ty}\;\UU         & \reduce \UU \\
[ \pi_1\,\sigma ]_{\Ty}\;\UU     & \reduce \UU,
\end{alignat*}
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.

@@

Yet, as we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent.
This boils down to that $[\blank]_{\Ty}\blank$ preserves all equality constructors.
\LT[noinline]{Should we mention any general condition for coherence?} 
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold (propositionally).
  The first two identities of the first row hold definitionally by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
For a QII definition \cite{Altenkirch2016a}, this extension adds following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}}
\end{alignat*}
as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.

In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes definitional. 
Likewise, the equality constructor $[]\El$ may be replaced by 
\begin{alignat*}{3}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma] {u})
\end{alignat*}
Yet, this naive change breaks the local confluence!
Observe that the local peak
\[
  \El\,u \Leftarrow [\idS]_{\Ty}(\El\,u) \reduce \El([\idS]\,u).
\]
cannot reduce to the same term, since the term substitution $[\idS]\,u$ is merely a constructor.
